Donkeys to Monoids: A Computational Framework for Dynamic Semantics

### 2013-05-19 paper.tex
Made Main.hs read from a file Added more appendices

\usepackage{stmaryrd}
\usepackage[utf8]{inputenc}
\usepackage{amsthm,amssymb,amsmath,dsfont,mathrsfs,nicefrac,wasysym,listings}
\usepackage{tikz}
\usepackage{enumitem}

@@

\lstloadlanguages{Haskell}
\lstnewenvironment{haskell}
    {\lstset{}%
      \csname lst@SetFirstLabel\endcsname}
    {\csname lst@SaveFirstLabel\endcsname}
    \lstset{
      basicstyle=\small\ttfamily,
      escapeinside={/+}{+/},
      flexiblecolumns=false,
      xleftmargin=0.05\textwidth,
      basewidth={0.5em,0.45em},
      literate={+}{{$+$}}1
               {/}{{$/$}}1
               {*}{{$*$}}1
               {=}{{$=$}}1
               {>}{{$>$}}1
               {<}{{$<$}}1
               {\\}{{$\lambda$}}1
               {++}{{$+\!\!\!+$}}1
               {::}{{$:\!\!\!:$}}1
               {\\\\}{{\char`\\\char`\\}}1
               {->}{{$\rightarrow$}}2
               {>=}{{$\geq$}}2
               {<-}{{$\leftarrow$}}2
               {<=}{{$\leq$}}2
               {=>}{{$\Rightarrow$}}2
               {/=}{{$\neq$}}2
               {.}{{$\circ$}}2
               {<<}{{$\ll$}}2
               {>>}{{$\gg$}}2
               {>>=}{{$\gg\!=$}}2
               {>=>}{{$>\!=\!>$}}2
               {|}{{$\mid$}}1
               {undefined}{{$\bot$}}1
               {`elem`}{{$\in$}}1
    }

@@

Donkey Monoids are a set of variations of Dynamic Predicate Logic (DPL) introduced by Albert Visser. The rationale behind them is a stride towards a logic that can more easily approximate natural language in aspects such as polarity and scope. Indeed their name derives from the famous `Donkey Sentence' by Geach\cite{geach1962reference} which exhibit these issues so well.

Visser didn't test his ideas computationally and neither did most people involved with DPL. In fact most of the classical literature on logification of natural language has this `problem'. In this paper I supply such an experiment, combining technologies introduced over the last four decades, creating a fully functioning `English to logic' framework.

The transformation rules of the supported English fragment are defined as a Montague Grammar, the semantics of which is discussed and defined in the first part of the paper. I apply the framework to a large collection of sentences, showing the conversion process step by step. Finally I discuss ways to enlarge the supported fragment of English, how we can combat and interprete the inherent ambiguity we meet, and how the flexibility of Donkey Monoids relate to the classical Montague lambda structures.

@@

\smallskip\smallskip
\noindent \small{\textbf{Keywords:} Montague Grammar, Dynamic Predicate Logic, Natural Language, Meaning, Donkey, Monoid, Computational, Haskell}

@@

\section{Introduction (5)}
% 5 pagish Donkey example to follow through.

@@

There are many reasons why it is interesting to represent information in a logical language. For a computer scientist, the biggest advantage is that we can compute with logic. There is a horde of algorithms for doing querying, question answering, reasoning, unification and resolution. They key to this powerhouse is obtaining a first order logic form of your information.

@@

\item[Robots] Since the beginning of Artificial Intelligence, logic based approaches have done well.\cite{russell2010artificial} Robots need to store relationships they learn about the world and react to it. For modern `robots' like apples Siri to be able to communicate with people, a translation layer between natural language and logic is a plausible path.
\end{description}

@@

The reason logics like DPL and Visser Monoids exist is a famous sentence noted by Peter Thomas Geach in 1962\cite{geach1962reference}: `Every farmer who owns a donkey beats it'. As we will see, this sentence illustrates that even if we restrict ourselves to factual sentences, formal first order logic still runs into a lot of problems when representing many natural sentences. In fact this sentence broke early automatic approaches to logification, and led to the construction of Discourse Representation Theory (DRT)\cite{kamp1993discourse}.

@@

\item `All his donkeys are treated well': This sentence somewhat willingly let's itself translate into: $\forall y(his(x,y)\wedge donkey(y) \rightarrow treated\_well(y))$, which has a free variable $x$ referring to some (male?) entity in the context of the discourse.

\item `Every farmer beats his donkey': This sentence clearly has two possible meanings: $\forall x(farmer(x)\rightarrow \exists y (beats(x,y)\wedge his(x,y)\wedge donkey(y))$ if `his' refers to the farmers or $\exists y(his(x,y)\wedge donkey(y)\wedge \forall z(farmer(z) \rightarrow beats(z,y))$ if it refers to somebody else. There is no way we can know without more context information.

@@

These examples should hopefully have shown that some uses of English, especially example 4 and 5, are well beyond what we can hope to automatically logify. However you might think the sentences of `factual nature' like 1 and 2 shouldn't be too hard? Unfortunately, as sentences get longer and more complex we will see that they are, and this is where Donkey Monads, Dynamic Predicate Logic (DPL) and Discourse Representation Theory (DRT) have a role to play. Let's take a close look at that `Donkey Sentence': `If a farmer owns a donkey, he beats it'.

@@

What turns out to be the correct logification is this:
%
\begin{align}
\forall x (farmer(x) \rightarrow \forall y (owns(x,y) \wedge donkey(y) \rightarrow beats(x,y))) \label{correct_donkey}
\end{align}
%
A far cry from our original `guess'. Moreover the above version even have the oddity that the interpretation of `all' is `the logical one': The formula \eqref{correct_donkey} is, in addition to the intended situations, also considered true if no farmers exist in the domain, or none of them own donkeys.

Now the above example may seem a bit contrived. Why did I have to try to combine the meaning from the obviously wrong pieces? Wasn't some other strategy more obvious? Before we go into details of such logification strategies, let's first see how easy it could have been, if we instead of first order logic had used Visser's Donkey Monoids:
%
\begin{align}
\Bowtie \cdot \exists x \cdot farmer(x) \cdot owns(x,y) \cdot\triangle\cdot\exists y \cdot donkey(y) \cdot\triangle\cdot\Bowtie\cdot beats(x,y)
\end{align}
%
If we read $\exists$ as `a' this is basically identical to the English sentence. The donkey monoid even supports placing the transitive verb between the two referants. Clearly some magic is going on, that we will have to investigate further.

Neat as the Donkey Monoids are, interested as we are in automation, we should be very concerned about those magical $\Bowtie$s and $\triangle$s that have popped up in our formula. What do they mean? How do they work? And how do we know when to use them? Visser himself, while answering question 1 and 2 in detail, doesn't give many hints on question 3. Only that they are probably controlled by adverbs and similar `helping words'.

@@

In the early sixties the American logician, Richard Montague, was a teaching a class in which the students had to do logification of simple sentences. He noticed that there were many patterns in the way they did it, and conjectured that it would be possible to formalize the entire process. So he set out to write a series of papers that became very influential in the philosophy of language, linguistics and computer science.

The Montague Grammar, as Montague's approach came to be know, works amazingly well, it's simplicity taken into account. It is not hard to see how one might convince oneself that it could be used to describe all of language. The basic idea is to define logical formulas recursively over the so called constituent tree of a sentence. The tree is based on the idea of context free grammars for natural language developed mainly by Noam Chomsky\cite{chomsky2002syntactic}. Figure \ref{fig:const_tree} is an example of what it might look like.

@@

The main thing to notice now is that for Montague's approach to work, it is crucial that we can always get the meaning of a composed phrase from a composition of the meaning of its subphrases. This compositional property is exactly what we saw earlier that first order logic doesn't posses, and Montague goes through many hoops to simmulate it.

I do however avoid many of those problems, since my goal here is not to test the validity of Montague's proposed rules, but to experiment with how Donkey Monoids and to an extent DPL behaves when put in the framework described. To draw a full circle I am however then going to translate the Donkey Monoids into first order logic so we can easily check if our formulas have the meanings we expect.

@@

\item[2 POS tagger] This is the first of three Stanford NLP programs my Haskell script is going to pass the text to. The POS tagger tokenizes the sentences and statstically tries to figures out part of speach for each word. This is our first source of error.

@@

\item[4 Correference Resolution] A lot have been said about how anaphora might be resolved using formal Montague Grammar like approaches. DRT is especially focused on this topic. None the less the current champeon of anaphora competitions\cite{lee2011stanford}\cite{raghunathan2010multi} is the heuristical Stanford Deterministic Coreference Resolution System, the workings of which I will explain in detail.

@@

\item[6 Dynamic Predicate Logic] Since the Donkey Monoids are variants of DPL, they are easy to convert into it. It is much more surprising that DPL lets itself convert into first order logic.

\item[7 First Order Logic] We end up translating our dynamic logics into this old friend. This is mainly done to test the conversion algorithm, and because the first order logic is what we are used to read. The results can be surprising.
\end{description}

@@

\section{Formal Stuff (10)}

@@

In the previous chapter I showed how First Order Logic lack certain compositional properties. A possible solution to this problem is to use a so called `update logic' (a logic with update semantics). Update semantics are logics inspired on the semantics of programming languages where a variable, once introduced, will stick around for a period not know at the time of it's introduction.

The idea of those logics is a `growth of information in time'. Starting from no information, and with each statement adding more. Information is here seen in the information theoretic way, where no information is when everything is possible ($\top$) and total information is when nothing is possible ($\bot$).

@@

Notice here in particular the interesting the `dependent' arrow. This arrow communicates that there is a one-many relationship between farmer and donkey, and hence that we are talking about a specific donkey for each farmer, not a single poor donkey beaten by everyone.

@@

DPL also doesn't provide any guidance into resolving anaphora. Just like giving meaning to the actual concepts, this must be handled outside of the logic. We will see later in this paper that this stance makes sense, since the current state of the art anaphora resolvers are simply based on simple heuristics over noun phrases.

In the end of this chapter we will see how a certain monadic construction by Visser can give us as much flexibility in DPL as we would get with a more involved system such as DRT or FCS.
\end{itemize}

@@

Dynamic Predicate Logic has been developed by a wide group of people: Jan van Eijck and Fer-Jan de Vries\cite{eijck1992dynamic}, Hans Kamp, Veltman, Groenendijk and Stokhof. It is the linguist version of Dynamic Logic invented by Pratt\cite{pratt1976semantical} which is again build on top of Floyd-Hoare Logic for reasoning about computer programs.

@@

Talking about binding, also notice that our union or `or' construct is a bit special. $\psi_1$ and $\psi_2$ cannot bind variables in each other, but in $(\psi_1\cup\psi_2)\cdot\psi_3$ they can both bind variables in $\psi_3$. This is going to introduce a lot of weird situations later. In the next section we will look closer at these problems, and also consider the alternative definition of `or': $\neg(\neg\psi_1\cdot\neg\psi_2)$.

\subsubsection{The problem with disjunction}

We won't actually use `or' in the implementation.
Example of the problems we still have in the logic.
Can't tell what variables are available where.

A major problem with our use of `relation union' as disjunction is the way it allows seemingly sound sentences to combine into syntactically invalid ones. Consider the sentence `Either a farmer owns a donkey or he doesn't'. We would write that as
\begin{align}
\exists x\cdot farmer(x)\cdot(\exists y\cdot owns(x,y)\cdot donkey(y)\vee \bot) \label{owns_donkey_or_not}
\end{align}
Now concatenating this sentence with `he beats it': $beat(x,y)$ would give us a problem. In fact this problem would be more than a syntactic problem, because before we know the model and the existence of the donkey, we don't know if the sentence is valid or not.

There is a cure to this problem. When we extend our logic with more power in the monad chapter, we get a second chance. However Let's first have a look at an obvious alternative definition for disjunction considered by Groenendijk\cite{groenendijk1991dynamic}. We'll call it `static disjunction':
\begin{align}
\alpha[\psi_1\vee\psi_2]\beta
 :\equiv&\ \alpha[\neg(\neg\psi_1\wedge\neg\psi_2)]\beta \nonumber\\
 \leftrightarrow&\ \alpha[\neg\psi_1\rightarrow\psi_2]\beta \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha[\neg\psi_1]\gamma\rightarrow\gamma\models[\psi_2]) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma((\alpha=\gamma\wedge\gamma\not\models\psi_1)\rightarrow\gamma\models[\psi_2]) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha=\gamma\rightarrow(\gamma\models\psi_1\vee\gamma\models\psi_2)) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge (\alpha\models\psi_1\vee\alpha\models\psi_2) \label{sem_or}
\end{align}

\subsubsection{Connection with first order logic}

@@

And combining the two monads we can do our original sentence as
%
\begin{quote}
If a farmer owns a donkey, he beats it
\end{quote}
%
as
\begin{equation}
\Bowtie \cdot \triangle \cdot \exists x \cdot farmer(x) \cdot \triangle \cdot owns(x,y) \cdot \triangle \cdot \exists y \cdot donkey(y) \cdot \triangle \cdot \Bowtie \cdot beats(x,y)
\end{equation}
%
In stream form, that is:
%
\begin{equation}
\begin{tabular}{|cllll|}
    \hline
    $(-,1)$ & $\exists x \cdot farmer(x)$ & ~ & ~ $ \exists y \cdot donkey(y)$ & ~ \\\hline
    $(-,0)$ & ~ & $owns(x,y)$ & ~ & ~ \\\hline
    $(+,1)$ & ~ & ~ & ~ & ~ \\\hline
    $(+,0)$ & ~ & ~ & ~ & $beats(x,y)$ \\\hline
\end{tabular}
\end{equation}
%
With the meaning being $(-,1) \cdot (-,0) \rightarrow (+,1) \cdot (+,0)$. Notice that the negative relations are again put on the left side of the implication, and the relations with larger scope value are put before those of lower value.

@@

For now we define the monad as follows:
%
\begin{flalign}
&\top := \langle \top, \top, \top, \top, +, 0 \rangle & \\
&\bot := \langle \top, \top, \top, \bot, +, 0 \rangle & \\
&\Bowtie := \langle \top, \top, \top, \top -, 0 \rangle & \\
&\triangle := \langle \top, \top, \top, \top +, 1 \rangle & \\
&\langle q_{-,1}, q_{-,0}, q_{+,1}, q_{+,0}, \alpha, i \rangle \cdot \langle r_{-,1}, r_{-,0}, r_{+,1}, r_{+,0}, \beta, j \rangle := &\\
&\hspace{1cm}\langle q_{-,1} \cdot r_{-\alpha,1+i}, q_{-,0} \cdot r_{-\alpha,i}, q_{+,1} \cdot r_{\alpha,1+i}, q_{+,0} \cdot r_{\alpha,i}, \alpha\beta, i+j \rangle& \nonumber
\end{flalign}

@@

While the above remains a hot topic in the philosophy of language, we are going to take a more practical stand. In our quest to automate the translation of language into Visser's monadic DPL, it is interesting that Montague was able to define such a large fragment of English using familiar tools of compositionality and lambda calculus.

@@

\subsubsection{Weak and strong readings}

@@

The Stanford procedure above is currently the strongest competitor in various coreference competitions.

\section{Implementation (8)}

@@

\begin{haskell}
data Fol = T | F | ,,,
type IFol = Interpretation -> Bool

data Dpl = DT | DF | ,,,
type IDpl = Interpretation -> Assignment -> Assignment -> Bool
type IDpl2 = Interpretation -> Assignment -> [Assignment]

data Visser1 = VT | VF | ,,,
type IVisser1 = (Dpl, Dpl, Integer)
data Visser2 = VT | VF | ,,,
type IVisser2 = (Dpl, Dpl, Dpl, Dpl, Integer, Integer)
\end{haskell}

@@

\subsection{The grammar}

@@

\section{Conclusions and Perspectives (2)}

@@

\subsection{Personal Report*}

What the...

\begin{itemize}
\item Studied computational linguistics and learned about state of the art technologies
\item Found out what was going on with dynamic logics and 
similar discourse algebras
\item Ran into many issues some of which have been discussed in great deal before (like strong vs. weak readings) and some that haven't had as much attention (like semantics of `and' and `or').
\item Combined the technologies and theory in a way not tried before, which yielded very good results
\end{itemize}

@@

\newpage
\section{Appendix}
\subsection{Derivation of DPL preconditions}
Remember that we have defined, for assignments $\alpha$ and $\beta$, DPL formula $\psi$ and FOL formula $\phi$: $\alpha\models\langle\psi\rangle\phi$ to mean $\exists\beta(\alpha\llbracket\psi\rrbracket\beta\wedge\beta\models\phi)$. $\alpha\models\phi$ is just the usual $\models$ from FOL.

@@

\subsection{Source Code*}

Should I include this?

>>>>

\usepackage{stmaryrd}
\usepackage[utf8]{inputenc}
\usepackage{amsthm,amssymb,amsmath,dsfont,mathrsfs,nicefrac,wasysym,listings}
\usepackage{tikz}
\usepackage{enumitem}
\usepackage{courier}

@@

\lstdefinestyle{haskellstyle}{%
  basicstyle=\scriptsize\ttfamily,
  language=Haskell,
  escapeinside={/+}{+/},
  flexiblecolumns=false,
  xleftmargin=-0.05\textwidth,
  basewidth={0.5em,0.45em},
  breaklines=true,
  literate={+}{{$+$}}1
           {/}{{$/$}}1
           {*}{{$*$}}1
           {=}{{$=$}}1
           {>}{{$>$}}1
           {<}{{$<$}}1
           {\\}{{$\lambda$}}1
           {++}{{$+\!\!\!+$}}1
           {::}{{$:\!\!\!:$}}1
           {\\\\}{{\char`\\\char`\\}}1
           {->}{{$\rightarrow$}}2
           {>=}{{$\geq$}}2
           {<-}{{$\leftarrow$}}2
           {<=}{{$\leq$}}2
           {=>}{{$\Rightarrow$}}2
           {/=}{{$\neq$}}2
%           { . }{{$\circ$}}2
           {<<}{{$\ll$}}2
           {>>}{{$\gg$}}2
           {>>=}{{$\gg\!=$}}2
           {>=>}{{$>\!=\!>$}}2
           {|}{{$\mid$}}1
           {undefined}{{$\bot$}}1
           {`elem`}{{$\in$}}1
           {⋈}{{$\Bowtie$}}1
           {⋅}{{$\cdot$}}1
           {△}{{$\triangle$}}1
           {∀}{{$\forall$}}1
           {∃}{{$\exists$}}1
           {⊥}{{$\bot$}}1
           {⊤}{{$\top$}}1
           {→}{{$\Rightarrow$}}1
           {∨}{{$\vee$}}1
           {¬}{{$\neg$}}1
           {∧}{{$\wedge$}}1
}

\lstdefinestyle{bashstyle}{%
  basicstyle=\scriptsize\ttfamily,
  escapeinside={/+}{+/},
  flexiblecolumns=false,
  xleftmargin=-0.05\textwidth,
  breaklines=true,
  basewidth={0.6em,0.45em},
  literate = {⋈}{{$\Bowtie$}}1
           {⋅}{{$\cdot$}}1
           {△}{{$\triangle$}}1
           {∀}{{$\forall$}}1
           {∃}{{$\exists$}}1
           {⊥}{{$\bot$}}1
           {⊤}{{$\top$}}1
           {→}{{$\Rightarrow$}}1
           {∨}{{$\vee$}}1
           {¬}{{$\neg$}}1
           {∧}{{$\wedge$}}1
}

\lstnewenvironment{haskell}{\lstset{style=haskellstyle}}{}

@@

In this paper I build a fully automatic system for translation of natural language into formal logic. I combine a variety of tools and theories to target standard issues such as anaphora resolution, sentence polarity etc. In particular I base my framework on dynamic logics in form of Dynamic Predicate Logic and a set of enhancements by Albert Visser\cite{visser1999donkey}.

`Donkey Monoids', as I will call these monoidal DPL variants, are part of a movement in linguistics towards logics that can more naturally approximate human language in aspects such as polarity and scope. Indeed their name derives from the famous `Donkey Sentence' by Geach\cite{geach1962reference} which exhibit these issues so well.

Visser didn't test his ideas computationally and neither did most people involved with DPL. In fact most of the classical literature on logification of natural language exhibits this `problem': Not testing computationally makes it easy to miss obvious details. With the present paper I supply such an experiment.

I build a rewriting system with a well defined set rules, discussing possible semantics. I apply the framework to a large collection of sentences, showing the conversion process step by step. Finally I discuss ways to enlarge the supported fragment of English, how we can combat and interprete the inherent ambiguity we meet, and how the flexibility of Donkey Monoids relate to the classical Montague lambda structures.

@@

\smallskip\smallskip
\noindent \small{\textbf{Keywords:} Natural Language Semantics, Meaning, Dynamic Predicate Logic, Donkey, Monoid, Montague Grammar, Computational, Haskell}

@@

\section{Introduction}
% 5 pagish Donkey example to follow through.

@@

There are many reasons why it is interesting to represent information in a logical language. For a computer scientist, the biggest advantage is that we can compute with logic. There is a horde of algorithms for doing querying, question answering, reasoning, unification and resolution - to all of which they key is a formal logic definition of your information.

@@

\item[Robots] Since the beginning of Artificial Intelligence, logic based approaches have done well.\cite{russell2010artificial} Robots need to store relationships they learn about the world and react to it. For modern `robots' like Apple's Siri to be able to communicate with people, a translation layer between natural language and logic is a plausible path.
\end{description}

@@

The reason logics like DPL and Visser Monoids exist is a famous sentence noted by Peter Thomas Geach in 1962\cite{geach1962reference}: `Every farmer who owns a donkey beats it'. As we will see, this sentence illustrates that even if we restrict ourselves to factual sentences, formal first-order logic still runs into a lot of problems when representing many natural sentences. In fact this sentence broke early automatic approaches to logification, and led to the construction of Discourse Representation Theory (DRT)\cite{kamp1993discourse}.

@@

\item `All his donkeys are treated well': This sentence somewhat willingly lets itself translate into: $\forall y(his(x,y)\wedge donkey(y) \rightarrow treated\_well(y))$, which has a free variable $x$ referring to some (male?) entity in the context of the discourse.

\item `Every farmer beats his donkey': This sentence clearly has two possible meanings: $\forall x(farmer(x)\rightarrow \exists y (beats(x,y)\wedge his(x,y)\wedge donkey(y))$, if `his' refers to the farmers, or $\exists y(his(x,y)\wedge donkey(y)\wedge \forall z(farmer(z) \rightarrow beats(z,y))$ if it refers to somebody else. There is no way we can know without more context information.

@@

Hopefully those examples make a convincing argument that some uses of English, especially example 4 and 5, are well beyond what we can hope to automatically logify. However you might think the sentences of `factual nature' like 1 and 2 look pretty trivial? In this paper we will see that they are certainly possible, but as sentences get longer and more complex, certainly tricky, and this is where Donkey Monoids, Dynamic Predicate Logic (DPL) and Discourse Representation Theory (DRT) come into play. Let's take a close look at that `Donkey Sentence': `If a farmer owns a donkey, he beats it'.

@@

What turns out to be the correct logification is this:
%
\begin{align}
\forall x (farmer(x) \rightarrow \forall y (owns(x,y) \wedge donkey(y) \rightarrow beats(x,y))) \label{correct_donkey}
\end{align}
%
A far cry from our original `guess'. Moreover the above version even has the oddity that the interpretation of `all' is `the logical one': The formula \eqref{correct_donkey} is, in addition to the intuitive situations\footnote{For non logicians at least.}, also considered true if no farmers exist in the domain, or none of them own donkeys.

Now the above example may seem a bit contrived. Why did I have to try to combine the meaning from the obviously wrong pieces? Wasn't some other strategy more obvious? Before we go into details of such logification strategies, let's first see how easy it could have been, if we instead of first-order logic had used Visser's Donkey Monoids:
%
\begin{align}
\Bowtie \cdot \exists x \cdot farmer(x) \cdot owns(x,y) \cdot\triangle\cdot\exists y \cdot donkey(y) \cdot\triangle\cdot\Bowtie\cdot beats(x,y)
\end{align}
%
If we read $\exists$ as `a' this is basically identical to the English sentence. The donkey monoid even supports placing the transitive verb between the two referants. Clearly some magic is going on, that we will have to investigate further.

Neat as the Donkey Monoids are, and interested as we are in their automation, we should be very concerned about those magical $\Bowtie$s and $\triangle$s which have popped up in our formula. We'll want to ask questions such ask (1) What do they mean? (2) How do they work? And (3) how do we know when to use them? Visser himself, while answering question 1 and 2 in great detail, doesn't give many hints on 3. Only that they `switches' are probably controlled by adverbs and other support words.

@@

In the early sixties the American logician, Richard Montague, was teaching a class in which the students had to do logification of simple sentences. He noticed that there were many patterns in the way they did it, and conjectured that it would be possible to formalize the entire process. So he set out to write a series of papers that became very influential in the philosophy of language, linguistics and computer science.

The Montague Grammar, as Montague's approach came to be know, has survived in computational linguistics for decades. Perhaps its main charm is in its rich promises for a simple, consistent way to describe all of language. The basic idea is to define logical formulas recursively over the so called constituent tree of a sentence. The tree is based on the idea of context free grammars for natural language developed mainly by Noam Chomsky\cite{chomsky2002syntactic}. Figure \ref{fig:const_tree} is an example of what one might look like.

@@

The main thing to notice now is that for Montague's approach to work, it is crucial that we can always get the meaning of a composed phrase from a composition of the meaning of its subphrases. This compositional property is exactly what we saw earlier that first-order logic doesn't posses, and Montague goes through many hoops to simulate it.

I do however avoid many of those problems, since my choices of logic behave much more compositionally than the FOL in Montague's proposed rules. To draw a full circle I am however going to translate the obtained Donkey Monoid representations into first-order logic so we can easily check if our formulas have the meanings we expect.

@@

\item[2 POS tagger] This is the first of three Stanford NLP programs my Haskell script is going to pass the text to. Then the POS tagger tokenizes the sentences and statstically tries to figures out part of speech for each word. This is also my first source of error.

@@

\item[4 Correference Resolution] A lot has been said about how anaphora might be resolved using formal Montague Grammar like approaches. DRT is especially focused on this topic. None the less the current champion of anaphora competitions\cite{lee2011stanford}\cite{raghunathan2010multi} is the heuristical Stanford Deterministic Coreference Resolution System, the workings of which I will explain in detail.

@@

\item[6 Dynamic Predicate Logic] Since the Donkey Monoids are variants of DPL, they are easy to convert into it. It is much more surprising that DPL lets itself convert into first-order logic.

\item[7 First-Order Logic] We end up translating our dynamic logics into this old friend. This is mainly done to test the conversion algorithm, and because the first-order logic is what we are used to read. The results can be surprising.
\end{description}

@@

\section{Dynamic Semantics and Montague}

@@

In the previous chapter I showed how first-order logic lacks certain compositional properties. A possible solution to this problem is to use a so called `update logic' (a logic with update semantics). Update semantics are logics inspired on the semantics of programming languages where a variable, once introduced, will stick around for a period not know at the time of it's introduction.

The idea of those logics is a `growth of information in time', starting from no information, and with each statement adding more. Information is here seen in the semantic way, where no information is when everything is possible ($\top$) and total information is when nothing is possible ($\bot$).

@@

Notice here in particular the interesting `dependent' arrow. This arrow communicates that there is a one-many relationship between farmer and donkey, and hence that we are talking about a specific donkey for each farmer, not a single poor donkey beaten by everyone.

@@

DPL also doesn't provide any guidance into resolving anaphora. Just like giving meaning to the actual concepts, this must be handled outside of the logic. We will see later in this paper that this stance makes sense, since the current state of the art anaphora resolvers are based on simple heuristics over noun phrases.

In the end of this chapter we will see how a certain monoid construction by Visser can give us as much flexibility in DPL as we would get with a more involved system such as DRT or FCS.
\end{itemize}

@@

Dynamic Predicate Logic has been developed by a wide group of people: Jan van Eijck and Fer-Jan de Vries\cite{eijck1992dynamic}, Hans Kamp, Veltman, Groenendijk and Stokhof. It is the linguist version of Dynamic Logic invented by Pratt\cite{pratt1976semantical} which is again built on top of Floyd-Hoare Logic for reasoning about computer programs.

@@

Talking about binding, also notice that our union or `or' construct is a bit special. $\psi_1$ and $\psi_2$ cannot bind variables in each other, but in $(\psi_1\cup\psi_2)\cdot\psi_3$ they can both bind variables in $\psi_3$. We won't actually use `or' in the main implementation\footnote{Turns out the Stanford tools find it weird as well.}, but in the appendix~\ref{sec:problem_disj} I look more closely at the weird situations that can arrise. I also consider the alternative definition of `or': $\neg(\neg\psi_1\cdot\neg\psi_2)$.

\subsubsection{Connection with First-Order Logic}

@@

And combining the two monoids we can do our original sentence as
%
\begin{quote}
If a farmer owns a donkey, he beats it
\end{quote}
%
as
\begin{equation}
\Bowtie \cdot \triangle \cdot \exists x \cdot farmer(x) \cdot \triangle \cdot owns(x,y) \cdot \triangle \cdot \exists y \cdot donkey(y) \cdot \triangle \cdot \Bowtie \cdot beats(x,y)
\end{equation}
%
In stream form, that is:
%
\begin{equation}
\begin{tabular}{|cllll|}
    \hline
    $(-,1)$ & $\exists x \cdot farmer(x)$ & ~ & ~ $ \exists y \cdot donkey(y)$ & ~ \\\hline
    $(-,0)$ & ~ & $owns(x,y)$ & ~ & ~ \\\hline
    $(+,1)$ & ~ & ~ & ~ & ~ \\\hline
    $(+,0)$ & ~ & ~ & ~ & $beats(x,y)$ \\\hline
\end{tabular}
\end{equation}
%
With the meaning being $(-,1) \cdot (-,0) \rightarrow (+,1) \cdot (+,0)$. Notice that the negative relations are again put on the left side of the implication, and the relations with larger scope value are put before those of lower value.

@@

For now we define the monoid as follows:
%
\begin{flalign}
&\top := \langle \top, \top, \top, \top, +, 0 \rangle & \\
&\bot := \langle \top, \top, \top, \bot, +, 0 \rangle & \\
&\Bowtie := \langle \top, \top, \top, \top -, 0 \rangle & \\
&\triangle := \langle \top, \top, \top, \top +, 1 \rangle & \\
&\langle q_{-,1}, q_{-,0}, q_{+,1}, q_{+,0}, \alpha, i \rangle \cdot \langle r_{-,1}, r_{-,0}, r_{+,1}, r_{+,0}, \beta, j \rangle := &\\
&\hspace{1cm}\langle q_{-,1} \cdot r_{-\alpha,1+i}, q_{-,0} \cdot r_{-\alpha,i}, q_{+,1} \cdot r_{\alpha,1+i}, q_{+,0} \cdot r_{\alpha,i}, \alpha\beta, i+j \rangle& \nonumber
\end{flalign}

@@

While the above remains a hot topic in the philosophy of language, we are going to take a more practical stand. In our quest to automate the translation of language into Visser's DPL monoid, it is interesting that Montague was able to define such a large fragment of English using familiar tools of compositionality and lambda calculus.

@@

\subsubsection{Weak and Strong Readings}

@@

The Stanford procedure above is currently the strongest competitor in various coreference competitions.\cite{lee2011stanford}\cite{raghunathan2010multi}

\section{Implementation}

@@

\begin{haskell}
data Fol = T | F | ...
type IFol = Interpretation -> Bool

data Dpl = DT | DF | ...
type IDpl = Interpretation -> Assignment -> Assignment -> Bool
type IDpl2 = Interpretation -> Assignment -> [Assignment]

data Visser1 = VT | VF | ...
type IVisser1 = (Dpl, Dpl, Integer)
data Visser2 = VT | VF | ...
type IVisser2 = (Dpl, Dpl, Dpl, Dpl, Integer, Integer)
\end{haskell}

@@

\subsection{The Grammar}

@@

\section{Conclusions}

@@

@@

\newpage
\section{Appendix}

\subsection{Derivation of DPL Preconditions}
Remember that we have defined, for assignments $\alpha$ and $\beta$, DPL formula $\psi$ and FOL formula $\phi$: $\alpha\models\langle\psi\rangle\phi$ to mean $\exists\beta(\alpha\llbracket\psi\rrbracket\beta\wedge\beta\models\phi)$. $\alpha\models\phi$ is just the usual $\models$ from FOL.

@@

\subsection{The Problem with Disjunction}\label{sec:problem_disj}

%We won't actually use `or' in the implementation.
%Example of the problems we still have in the logic.
%Can't tell what variables are available where.

A major problem with our use of the union of relations as disjunction is the way it allows seemingly sound sentences to combine into syntactically invalid ones. Consider the sentence `Either a farmer owns a donkey or he doesn't'. We would write that as
\begin{align}
\exists x\cdot farmer(x)\cdot(\exists y\cdot owns(x,y)\cdot donkey(y)\vee \bot) \label{owns_donkey_or_not}
\end{align}
Now concatenating this sentence with `he beats it': $beat(x,y)$ would give us a problem. In fact this problem would be more than a syntactic problem, because before we know the model and the existence of the donkey, we don't know if the sentence is valid or not.

There is a cure to this problem. When we extend our logic with more power in the monoid chapter, we get a second chance. However Let's first have a look at an obvious alternative definition for disjunction considered by Groenendijk\cite{groenendijk1991dynamic}. We'll call it `static disjunction':
\begin{align}
\alpha[\psi_1\vee\psi_2]\beta
 :\equiv&\ \alpha[\neg(\neg\psi_1\wedge\neg\psi_2)]\beta \nonumber\\
 \leftrightarrow&\ \alpha[\neg\psi_1\rightarrow\psi_2]\beta \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha[\neg\psi_1]\gamma\rightarrow\gamma\models[\psi_2]) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma((\alpha=\gamma\wedge\gamma\not\models\psi_1)\rightarrow\gamma\models[\psi_2]) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha=\gamma\rightarrow(\gamma\models\psi_1\vee\gamma\models\psi_2)) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge (\alpha\models\psi_1\vee\alpha\models\psi_2) \label{sem_or}
\end{align}

\subsection{Sample Output}

\lstinputlisting[style=bashstyle]{sample_output}

\subsection{Source Code}

\lstinputlisting[style=haskellstyle,caption=\lstname]{src/Main.hs}
\lstinputlisting[style=haskellstyle,caption=\lstname]{src/Stanford.hs}
\lstinputlisting[style=haskellstyle,caption=\lstname]{src/ToLogic.hs}
\lstinputlisting[style=haskellstyle,caption=\lstname]{src/Dpl.hs}
\lstinputlisting[style=haskellstyle,caption=\lstname]{src/Visser.hs}
\lstinputlisting[style=haskellstyle,caption=\lstname]{src/Tests.hs}
\lstinputlisting[style=haskellstyle,caption=\lstname]{src/Utils.hs}
